**Bayesian statistics for longitudinal studies in biomedical research**

### 2020-10-27 Manuscripts/Manuscript_AM_v1.Rmd
Updated manuscript

A longitudinal study is defined as that were a variable of interest is measured repeatedly in a group (or groups) of subjects. In biomedical research, this type of study is preferred when the intention is to observe the evolution of the effect of treatment (or treatments) across time, rather than analyzing the information at a single timepoint (a cross-sectional study). Clinical examples of this  approach in biomedical research include studies on breast and neck cancer[@sio2016;@kamstra2015], where in the first case  weekly measurements of skin toxicities in breast cancer patients with radiation-induced dermatitis were taken for up to 8 weeks, and mouth opening in head and neck cancer patients was measured at 6,12, 18, 24 and 36 months after radiotherapy (RT) in the latter. Similar examples are found in studies of tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression[@ritter2001;@roth2017], and cell metabolism[@jones2018;@skala2010]. From a statistical standpoint, a longitudinal study presents advantages over a cross-sectional approach:it requires a lower number of subjects to reach a certain statistical power, and besides the previously mentioned time-effect evolution, it allows to determine the intra-variability of the response between different subjects [@guo2013; @fitzmaurice2012].

Researchers have typically employed a _frequentist_ approach to analyze longitudinal data. Such type of analysis is based on a hypothesis test using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA). However, a rm-ANOVA analysis not only assumes that the data fulfills certain requisites such as constant variance across measurements (which is frequently unjustified) and  complete observations from each subject [@schober2018;@gueorguieva2004], but it also requires the use of _post hoc_ analyses which inflate the false positivity rate[@liu2010]. Recently, _linear mixed effects models_ (LMEMs) have started to be used by certain groups to analyze longitudinal data [@vishwanath2009;@skala2010]. Generally speaking, LMEMs incorporate both _fixed_ and _random_ effects making them more flexible that rm-ANOVA as they can work with missing observations, and they allow to model the covariance of the parameters in different manners[@west2014].

@@

* Section 1: Challenges presented by longitudinal studies: Missing observations, and correlation between measurements. How rm-ANOVA is limited by missing observations, and how both rm-ANOVA and LMEMs are limited with data that does not follow a linear trend (equations for both situations and the fit they produce.)

>>>>

A longitudinal study is defined as one where a variable of interest is measured repeatedly in a group (or groups) of subjects. In biomedical research, this type of study is preferred when the intention is to observe the evolution of the effect of treatment across time, rather than analyzing the information at a single timepoint (a cross-sectional study). Clinical examples of this  approach in biomedical research include studies on breast and neck cancer[@sio2016;@kamstra2015]; in the first case, weekly measurements of skin toxicities in  patients with radiation-induced dermatitis were taken for up to 8 weeks; whereas in the latter mouth opening measured was assessed at 6,12, 18, 24 and 36 months after radiotherapy (RT). Longitudinal studies have used also to measure tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression[@ritter2001;@roth2017], and cell metabolism[@jones2018;@skala2010]. From a statistical standpoint, a longitudinal study presents advantages over a cross-sectional approach:it requires a lower number of subjects to reach a certain statistical power, and besides it being able to track the previously mentioned time-effect evolution on a group-by-group basis, it allows to determine the variability of the response within subjects [@guo2013; @fitzmaurice2012]. In other words,a longitudinal study permits to quantify how the variable changes within each subject across time.

Researchers have typically employed a _frequentist_ approach to analyze longitudinal data. Such statistical view derives its name from the fact that it views probability as a limiting frequency [wagenmakers2008]. In practice, this analysis for longitudinal data is  based on a null hypothesis test using the _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA). However, rm-ANOVA makes two key assumptions regarding the data: constant correlation across same-subject measurements and  complete observations from each subject [@schober2018;@gueorguieva2004]. The first condition is frequently unjustified in many cases as  the correlation between measures diminishes when the time gap between them increases [@ugrinowitsch2004], and in that case violating that assumption increases the type I error rate[@lane2016]. The reasons behind the second assumption being not met are varied: subjects can withdraw during the course of the study, attrition in animals due to injury or weight loss, or complications can arise that prevent the researcher from collecting measurements at a required timepoint. When incomplete observations  exist, rm-ANOVA requires to exclude the subject with missing observations from the analysis, which can result in increased costs for the study and possibly bias the results.

Moreover, rm-ANOVA uses a _post hoc_ analysis to distinguish the differences between groups. Because a _post hoc_ analysis is based in multiple repeated comparisons, it can inflate the false positivity rate[@liu2010]. Recently, _linear mixed effects models_ (LMEMs) have started to be used by certain groups to analyze longitudinal data [@vishwanath2009;@skala2010]. Generally speaking, LMEMs incorporate both _fixed_ and _random_ effects making them more flexible that rm-ANOVA as they can work with missing observations, and they allow to model the covariance of the parameters in different manners[@west2014].

@@

* Section 1:

###Challenges presented by longitudinal studies:

Missing observations, and correlation between measurements. How rm-ANOVA is limited by missing observations, and how both rm-ANOVA and LMEMs are limited with data that does not follow a linear trend (equations for both situations and the fit they produce.)
